---
title: "countland: large sponge dataset"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{countland: pig and placozoa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Samuel H. Church

This document applies `countland` to the analysis of real-world scRNA-seq data.

The following packages are required to run the analysis:

```{r load,results=F,message=F}
library(countland)
library(Seurat)
library(glmpca)

library(knitr)
library(dplyr)
library(gridExtra)

library(ggplot2)
library(viridis)
color_palette <- c("#8c564b", "#9467bd", "#2ca02c", "#e377c2", "#d62728", "#17becf", "#bcbd22", "#ff7f0e", "#7f7f7f", "#1f77b4")
theme_set(theme_classic())
```

## Get the data

Here we apply countland to a dataset from a sponge (_Spongilla lacustris_) provided by [Musser _et al_ (2021)](https://www.science.org/doi/full/10.1126/science.abj2949).

We set up two matrices, one with all cells, and one with only differentiated cells.

```{r data}
sponge.data <- Seurat::Read10X(data.dir = "../data/sponge_all/")
sponge_diff.data <- Seurat::Read10X(data.dir = "../data/sponge_all_but_arch_or_trans/")
```

## Run countland

We then apply countland. Here we modify the standard countland workflow to filter the same genes and cells that are filtered with Seurat, standard workflow. We set the number of clusters to published number of cell labels.

```{r run,results=F}
run_filter_countland <- function(mtx){
  C <- countland(mtx,remove_empty = TRUE)
  
  C <- ScoreCells(C)
  cell_filt <- C@cell_scores[C@cell_scores$n_features >= 200,]$names
  C <- SubsetCells(C,which(C@names_cells %in% cell_filt),remove_empty=FALSE)
  
  C <- ScoreGenes(C,subsample=FALSE)
  gene_filt <- C@gene_scores[C@gene_scores$n_cells >= 3,]$names
  C <- SubsetGenes(C,which(C@names_genes %in% gene_filt),remove_empty=FALSE)
  
  return(C)
}

sponge_C <- run_filter_countland(sponge.data)
sponge_diff_C <- run_filter_countland(sponge_diff.data)

run_countland <- function(C){
  real_labels <- sapply(strsplit(C@names_cells, "_"), function(v){return(v[2])})
  n_clusters <- length(unique(real_labels))

  C <- Dot(C,subsample=FALSE)
  C <- Embed(C,n_components=n_clusters)
  C <- Cluster(C, n_clusters=n_clusters)
}


# measure run time using countland
start = Sys.time()
sponge_C <- run_countland(sponge_C)
C_time <- Sys.time()-start

sponge_diff_C <- run_countland(sponge_diff_C)
```

## Run Seurat

We first apply Seurat using the default approach as listed in [this tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html).

```{r seurat-function}
run_seurat <- function(mtx){
  obj <- CreateSeuratObject(counts=mtx, project="spo", min.cells = 3, min.features = 200)
  obj <- NormalizeData(obj, normalization.method = "LogNormalize", scale.factor = 10000)
  all.genes <- rownames(obj)
  obj <- ScaleData(obj, features=all.genes)
  obj <- FindVariableFeatures(object=obj, mean.function = ExpMean, dispersion.function = LogVMR)
  obj <- RunPCA(obj, features=VariableFeatures(obj), ndims.print = 1:50, npcs = 100)
  obj <- FindNeighbors(obj, dims = 1:40)
  obj <- FindClusters(obj)
  obj <- RunTSNE(object = obj, dims = 1:40)
  return(obj)
}

# measure run time with Seurat
start = Sys.time()
sponge_obj <- run_seurat(sponge.data)
seurat_time <- Sys.time()-start

sponge_diff_obj <- run_seurat(sponge_diff.data)
```

## Run GLM-PCA

We also visualize the data using GLM-PCA. This analysis can take a long time.

```{r glmpca}
run_glmpca <- function(counts){
  return(glmpca(as(counts,"matrix"),2))
}

sponge_glm <- run_glmpca(sponge_C@counts)
sponge_diff_glm <- run_glmpca(sponge_diff_C@counts)

```

## Compare computational time.

```{r time-table}
time_table <- format(data.frame(C_time, seurat_time, glm_time));colnames(time_table) <- c("countland","Seurat")
kable(t(time_table),caption="computational time",col.names="comp. time")

n_cells <- length(sponge_C@names_cells)
time_table <- cbind(n_cells,time_table);colnames(time_table) <- c("num.cells","countland","Seurat")
write.table(time_table,file="sponge_time_table.tsv",sep="\t",row.names=F)
```

## Plot the results.

```{r plot-full, fig.height=4, fig.width=8}
mtx <- sponge.data
obj <- sponge_obj
glm <- sponge_glm
C <- sponge_C

seurat_loading_1 <- obj@reductions$tsne@cell.embeddings[,1]
seurat_loading_2 <- obj@reductions$tsne@cell.embeddings[,2]

glm_loading_1 <- glm$factors[,1]
glm_loading_2 <- glm$factors[,2]

countland_loading_1 <- C@embedding[,5]
countland_loading_2 <- C@embedding[,7]

total_counts <- C@cell_scores$total_counts
n_features_0 <- apply(C@counts,2,function(x){sum(x==0)})

pub_labels <- sapply(strsplit(colnames(mtx), "_"), function(v){return(v[2])})
n_clusters <- length(unique(pub_labels))
countland_clusters <- as.character(C@cluster_labels)

sgdf <- data.frame(seurat_loading_1,seurat_loading_2,
  countland_loading_1,countland_loading_2,
  pub_labels,countland_clusters,total_counts,n_features_0)

get_cluster_comb_df <- function(cl1,cl2){
  return(data.frame('source'=cl1,'target'=cl2) %>% group_by(source,target) %>% count())
}
# align colors
r <- get_cluster_comb_df(pub_labels,countland_clusters)
# arrange by cluster composition
r1 <- r %>% group_by(source) %>% arrange(desc(n)) %>% slice(1) %>% ungroup() %>% mutate(color = viridis::turbo(n=n()))
# source colors
s <- r1$color;names(s) <- r1$source
# target colors
t <- r1 %>% distinct(target,.keep_all=T) %>% pull(color);names(t) <- r1 %>% distinct(target,.keep_all=T) %>% pull(target)
# fill in colors for rest of target clusters
r2 <- r %>% filter(!target %in% names(t)) %>% ungroup() %>% distinct(target) %>% mutate(color = gray.colors(n=n()))
t2 <- r2$color;names(t2) <- r2$target
t <- c(t,t2)


g1 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = pub_labels)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("tSNE - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g2 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = countland_clusters)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("tSNE - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + guides(colour = guide_legend(override.aes = list(size=6))) + 
  theme(legend.position="none")

g3 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = pub_labels)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("GLMPCA - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g4 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = countland_clusters)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("GLMPCA - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g5 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = pub_labels)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("spectral - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g6 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = countland_clusters)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("spectral - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

pdf(file="../../figures_and_panels/panel_sponge_full.pdf",width=8,height=6)
grid.arrange(g1,g3,g5,g2,g4,g6,nrow=2)
dev.off()

gg1 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = total_counts)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("tSNE - total counts") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

gg2 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = n_features_0)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("tSNE - number of zeros") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + guides(colour = guide_legend(override.aes = list(size=6))) + 
  theme(legend.position="none")

gg3 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = total_counts)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("GLMPCA - total counts") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

gg4 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = n_features_0)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("GLMPCA - number of zeros") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

gg5 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = total_counts)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = s) + 
  ggtitle("spectral - total counts") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

gg6 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = n_features_0)) + 
  geom_point(size=0.1) +
  scale_color_manual(values = t) + 
  ggtitle("spectral - number of zeros") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

pdf(file="../../figures_and_panels/panel_sponge_full.pdf",width=10,height=6)
grid.arrange(g2,gg1,gg2,g4,gg3,gg4,g6,gg5,gg6,nrow=3)
dev.off()

pub_legend <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = pub_labels)) + geom_point() + 
  scale_color_manual(values = s) + 
  guides(colour = guide_legend(override.aes = list(size=5)))
count_legend <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = countland_clusters)) + geom_point() + 
  scale_color_manual(values = t) + 
  guides(colour = guide_legend(override.aes = list(size=5)))
pdf(file="../../figures_and_panels/panel_legend_full.pdf",width=4,height=8)
grid.arrange(pub_legend,count_legend,nrow=2)
dev.off()
```

```{r plot-differentiated, fig.height=4, fig.width=8}
mtx <- sponge_diff.data
obj <- sponge_diff_obj
glm <- sponge_diff_glm
C <- sponge_diff_C

seurat_loading_1 <- obj@reductions$tsne@cell.embeddings[,1]
seurat_loading_2 <- obj@reductions$tsne@cell.embeddings[,2]

glm_loading_1 <- glm$factors[,1]
glm_loading_2 <- glm$factors[,2]

countland_loading_1 <- C@embedding[,2]
countland_loading_2 <- C@embedding[,3]

pub_labels <- sapply(strsplit(colnames(mtx), "_"), function(v){return(v[2])})
n_clusters <- length(unique(pub_labels))
countland_clusters <- as.character(C@cluster_labels)

sgdf <- data.frame(seurat_loading_1,seurat_loading_2,
  countland_loading_1,countland_loading_2,
  pub_labels,countland_clusters)

# align colors
r <- get_cluster_comb_df(pub_labels,countland_clusters)
# arrange by cluster composition
r1 <- r %>% group_by(source) %>% arrange(desc(n)) %>% slice(1) %>% ungroup() %>% arrange(desc(source)) %>% mutate(color = rev(viridis::turbo(n=n())))
# source colors
s <- r1$color;names(s) <- r1$source
# target colors
t <- r1 %>% distinct(target,.keep_all=T) %>% pull(color);names(t) <- r1 %>% distinct(target,.keep_all=T) %>% pull(target)
# fill in colors for rest of target clusters
r2 <- r %>% filter(!target %in% names(t)) %>% ungroup() %>% distinct(target) %>% mutate(color = gray.colors(n=n()))
t2 <- r2$color;names(t2) <- r2$target
t <- c(t,t2)

g1 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = pub_labels)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = s) + 
  ggtitle("tSNE - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g2 <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = countland_clusters)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = t) + 
  ggtitle("tSNE - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g3 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = pub_labels)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = s) + 
  ggtitle("GLMPCA - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g4 <- ggplot(sgdf,aes(x = glm_loading_1, y = glm_loading_2, color = countland_clusters)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = t) + 
  ggtitle("GLMPCA - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g5 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = pub_labels)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = s) + 
  ggtitle("spectral - published labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

g6 <- ggplot(sgdf,aes(x = countland_loading_1, y = countland_loading_2, color = countland_clusters)) + 
  geom_point(size=0.25) +
  scale_color_manual(values = t) + 
  ggtitle("spectral - countland clusters") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank()) + 
  theme(legend.position="none")

pdf(file="../../figures_and_panels/panel_sponge_diff.pdf",width=8,height=6)
grid.arrange(g1,g3,g5,g2,g4,g6,nrow=2)
dev.off()

pub_legend <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = pub_labels)) + geom_point() + 
  scale_color_manual(values = s) + 
  guides(colour = guide_legend(override.aes = list(size=5), reverse=T))
count_legend <- ggplot(sgdf,aes(x = seurat_loading_1, y = seurat_loading_2, color = countland_clusters)) + geom_point() + 
  scale_color_manual(values = t) + 
  guides(colour = guide_legend(override.aes = list(size=5), reverse=T))
pdf(file="../../figures_and_panels/panel_legend_diff.pdf",width=4,height=10)
grid.arrange(pub_legend,count_legend,nrow=2)
dev.off()
```

