# countland: performance evaluation - Silver standard
### Samuel H. Church

This document applies `countland` to the analysis of a benchmark scRNA-seq dataset to test its performance.

The following packages are required to run the analysis:

```{r load,results=F,message=F}
library(countland)
library(Seurat)

library(aricode)
library(clevr)

library(knitr)
library(dplyr)
library(ggplot2)
library(viridis)
library(gridExtra)
theme_set(theme_classic())

set.seed(84095)
```

## Get the data

We have used the Silver standard dataset(3A) provided by [Freytag _et al_ (2018)](10.12688/f1000research.15809.2). This data consists of 4,300 cells that have reported labels corresponding to human blood cell lines.

```{r data}
silver.data <- Seurat::Read10X(data.dir = "../../data/Silver_Freytag2018/SilverDataset3a_Freytag2018")
```

## Analyze the Silver standard dataset 

We first alculated several gene expression scores to create alternative matrices that take into account expression heterogeneity.

```{r countland}
C <- countland(silver.data,remove_empty = TRUE)
C <- ScoreGenes(C,subsample=F)
```

```{r alternate-matrices}
# downsample genes to a maximum total counts equal to the number of cells
C_gene <- Subsample(C,gene_counts = ncol(C@counts))
C_gene_low <- Subsample(C,gene_counts = ncol(C@counts)/2)

# subset genes to bottom 50% of expressed genes
subset <- C@names_genes %in% (C@gene_scores %>% filter(total_counts < quantile(total_counts,0.95)) %>% pull(names))
C_subset <- SubsetGenes(C,gene_indices = subset)

# downsample cells to standard sequencing depth
C_cell <- Subsample(C,cell_counts = "min")
C_gene_cell <- Subsample(C,gene_counts = ncol(C@counts)/2,cell_counts="min")
```

We ran countland on each of these matrices, using 11 clusters, corresponding to the reported cell labels.

```{r run-countland, results=F}
run_countland <- function(obj,clusters,components,subsample){
  obj <- Dot(obj,subsample=subsample)
  obj <- Embed(obj)
  obj <- Cluster(obj,n_clusters=clusters,n_components=components)
  return(obj)
}

C <- run_countland(C,11,11,F)
C_gene <- run_countland(C_gene,11,11,T)
C_gene_low <- run_countland(C_gene_low,11,11,T)
C_subset <- run_countland(C_subset,11,11,F)
C_cell <- run_countland(C_cell,11,11,T)
C_gene_cell <- run_countland(C_gene_cell,11,11,T)
```

We evaluated our results using three measures: the adjusted rand index, normalized mutual information, and cluster homogeneity.

```{r results}
res <- list(C,C_gene,C_gene_low,C_subset,C_cell,C_gene_cell)
rand <- sapply(res,function(x){ARI(gsub("_.*","",x@names_cells),x@cluster_labels)})
nmi <- sapply(res,function(x){NMI(gsub("_.*","",x@names_cells),x@cluster_labels)})
homog <- sapply(res,function(x){homogeneity(gsub("_.*","",x@names_cells),x@cluster_labels)})

silver_results <- round(data.frame('adjusted rand index' = rand, 'norm. mutual info.' = nmi, "homogeneity" = homog, check.names=F),3)
rownames(silver_results) <- c("raw counts","subsampled genes to n cells","subsampled genes to 0.5 * n_cells","dropped top 5% of genes","standard sequencing depth across cells","subsampled genes and standard sequencing depth")
kable(t(silver_results),caption="cluster evaluation")
```

We compared our results to those obtained using `Seurat` on the same data. 

```{r run-seurat}
# this function runs Seurat using the standard workflow
# default parameters are used, with the exception of resolution, which can be adjusted
# and the option of using only highly variable features, vs using all features
run_seurat <- function(counts,resolution,find_variable){
  object <- CreateSeuratObject(counts = counts, project = "gold")
  object <- NormalizeData(object)
  all.genes <- rownames(object)
  object <- ScaleData(object, features = all.genes)
  if(find_variable == F){
    object <- RunPCA(object, features = all.genes, verbose=F)
  } else {
    object <- FindVariableFeatures(object, selection.method = "vst", nfeatures = 2000)
    object <- RunPCA(object, features = VariableFeatures(object = object), verbose=F)    
  }
  object <- FindNeighbors(object, dims = 1:10)
  object <- FindClusters(object, resolution = resolution)
  object <- RunUMAP(object, dims = 1:10)
  return(object)
}

seurat_silver <- run_seurat(C@counts,0.8,find_variable=T)
seurat_rand <- ARI(gsub("_.*","",seurat_silver$orig.ident),seurat_silver@meta.data$seurat_clusters)
seurat_nmi <- NMI(gsub("_.*","",seurat_silver$orig.ident),seurat_silver@meta.data$seurat_clusters)
seurat_homog <- homogeneity(gsub("_.*","",seurat_silver$orig.ident),seurat_silver@meta.data$seurat_clusters)

silver_standard_results <- rbind(silver_results,round(c(seurat_rand,seurat_nmi,seurat_homog),3))
rownames(silver_standard_results) <- c("countland, raw counts","subsampled genes to n cells","subsampled genes to 0.5 * n_cells","dropped top 5% of genes","standard sequencing depth across cells","subsampled genes and standard sequencing depth","Seurat")
kable(silver_standard_results)
save(silver_standard_results,file="performance_evaluation_silver_standard_results.RData")
```

Here we visualize our results.

```{r plot,fig.height=8,fig.width=12,fig.align="center"}
u1 <- C_subset@embedding[,2];u2 <- C_subset@embedding[,3]
real_labels <- gsub("_.*","",C_subset@names_cells)
countland_clusters <- as.character(C_subset@cluster_labels)
gdf <- data.frame("component_1" = u1, "component_2" = u2, "real_labels" = real_labels, "countland_clusters" = countland_clusters)

g1 <- ggplot(gdf,aes(x = component_1, y = component_2, color = real_labels)) + 
  geom_point(size=1) +
  scale_color_manual(values = viridis(option="D",12)) + 
  ggtitle("ground truth cell labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())

g2 <- ggplot(gdf,aes(x = component_1, y = component_2, color = countland_clusters)) + 
  geom_point(size=1) +
  scale_color_manual(values = viridis(option="C",11)) + 
  ggtitle("countland: subset top 5%") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())

u1 <- seurat_silver@reductions$umap@cell.embeddings[,1]
u2 <- seurat_silver@reductions$umap@cell.embeddings[,2]
real_labels <- gsub("_.*","",seurat_silver$orig.ident)
seurat_clusters <- as.character(seurat_silver@meta.data$seurat_clusters)
sgdf <- data.frame("component_1" = u1, "component_2" = u2, "real_labels" = real_labels, "seurat_clusters" = seurat_clusters)

g3 <- ggplot(sgdf,aes(x = component_1, y = component_2, color = real_labels)) + 
  geom_point(size=1) +
  scale_color_manual(values = viridis(option="D",12)) + 
  ggtitle("ground truth cell labels") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())

g4 <- ggplot(sgdf,aes(x = component_1, y = component_2, color = countland_clusters)) + 
  geom_point(size=1) +
  scale_color_manual(values = viridis(option="F",11)) + 
  ggtitle("Seurat") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())

grid.arrange(g1,g2,g3,g4,ncol=2)
```



