---
title: "countland: pig and placozoa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{countland: pig and placozoa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Samuel H. Church

This document applies `countland` to the analysis of real-world scRNA-seq data.

The following packages are required to run the analysis:

```{r load,results=F,message=F}
library(countland)
library(Seurat)
library(aricode)
library(clevr)

library(knitr)
library(dplyr)
library(gridExtra)

library(ggplot2)
library(viridis)
color_palette <- c("#8c564b", "#9467bd", "#2ca02c", "#e377c2", "#d62728", "#17becf", "#bcbd22", "#ff7f0e", "#7f7f7f", "#1f77b4")
theme_set(theme_classic())
```

## Get the data

Here we apply countland to a dataset from the placozoan (_Trichoplax adherens_) provided by [Sebé-Pedros _et al_ (2018)](https://doi.org/10.1038/s41559-018-0575-6), and a dataset from the pig (_Sus scrofa_) provided by [van Zyl _et al_ (2020](https://www.pnas.org/cgi/doi/10.1073/pnas.2001250117).

```{r data}
Tri.data <- Seurat::Read10X(data.dir = "../../data/Trichoplax_SebePedros2018/")
Sus.data <- Seurat::Read10X(data.dir = "../../data/Pig9_vanZyl2020_mtx/")
```

## Run countland

We first apply spectral embedding to the two datasets.

```{r run,results=F}
Sus <- countland(Sus.data,remove_empty = TRUE)
Sus <- Subsample(Sus,gene_counts = ncol(Sus@counts)*50,cell_counts = "min")
Sus <- Dot(Sus,subsample=TRUE)
Sus <- Embed(Sus,n_components=20)

Tri <- countland(Tri.data,remove_empty = TRUE)
Tri <- Subsample(Tri,gene_counts = ncol(Tri@counts)*50,cell_counts = "min")
Tri <- Dot(Tri,subsample=TRUE)
Tri <- Embed(Tri,n_components=20)
```
Because we have no _a priori_ hypothesis as to the number of clusters that may be present, we check to see if the eigengap heuristic for spectral clustering is informative.

```{r elbow-Sus}
PlotEigengap(Sus)
```

For _S. scrofa_, it appears as though the largest visible gap at the highest value is between 7 and 8, so we'll use 7 clusters.

```{r elbow-Tri}
PlotEigengap(Tri)
```

For _T. adherens_, it appears that the largest visible gap is between 11 and 12, so we'll use 11 clusters.


```{r cluster}
Sus <- Cluster(Sus,n_clusters=7,n_components=7)
Tri <- Cluster(Tri,n_clusters=11,n_components=11)
```

We can visualize the results using spectral embedding.

```{r plot-Tri,fig.height=4,fig.width=6,fig.align="center"}
embed <- Tri@embedding[,2:3]
embed <- setNames(data.frame(embed),paste("component_",seq_len(2),sep=""))

ggplot2::ggplot(embed,aes(x = component_1,y = component_2, color=as.character(Tri@cluster_labels))) +
geom_point(size=1) +
guides(color=guide_legend(title="cluster")) +
scale_color_manual(values=c(color_palette,"black"))
```

In the case of _S. scrofa_ we see that the first two variable components from sepctral embedding are not very informative, so it's helpful to view additional components.

```{r plot-Sus,fig.height=4,fig.width=6,fig.align="center"}
Scl <- paste0("countland_cluster:",as.character(Sus@cluster_labels))
Scolor <- viridis::turbo(n=8)
Sdf <- setNames(data.frame(Sus@embedding[,2:5]),paste0("component_",seq_len(4)))
Sdf$clusters <- Scl
g1 <- ggplot(Sdf,aes(x = component_1, y = component_2, color = clusters)) + 
  geom_point(size=1) + 
  scale_color_manual(values=Scolor) +
  theme("legend.position" = "none") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())
g2 <- ggplot(Sdf,aes(x = component_3, y = component_4, color = clusters)) + 
  geom_point(size=1) + 
  scale_color_manual(values=Scolor) +
  theme("legend.position" = "none") + 
  theme(axis.ticks = element_blank(),axis.text = element_blank())
grid.arrange(g1,g2,ncol=2)
```

